Digging Deeper Into the State of the Practice for Domain Specific Research Software

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
First complete draft - 1.5 pages too long

	To improve software development methods and tools for research software, we
	first need to understand the current state of the practice.  Therefore, we
	have developed a methodology for assessing the state of the software
	development practices for a given research software domain.  The methodology
	is applied to one domain at a time in recognition that software development in
	different domains is likely to have adopted different best practices.
	Moreover, providing a means to measure different domains facilitates
	comparison of development practices between domains.  For each domain we wish
	to answer questions such as: 
  \begin{inparaenum}[i)]
    \item What artifacts (documents, code, test cases, etc.) are present?
    \item What tools are used?
    \item What principles, process and methodologies are used?
    \item What are the pain points for developers?
    \item What actions are used to improve qualities like maintainability and
    reproducibility?
  \end{inparaenum} 
  To answer these questions, our methodology prescribes the following steps: 
  The abstract should briefly summarize the contents of the paper in 150--250 words.

\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}

@@

Close to our goal of 160 person hours, we estimate 173 hours to complete the
assessment of a given domain~\cite{SmithEtAl2021}.  This estimate assumes that
the domain has already been decided and the Domain Expert has been recruited.
The estimate assumes that the template spreadsheets and developed AHP
tool~\cite{SmithEtAl2021}, will be employed, rather than developing new tools.
The person hours given are a rough estimate, based on our experience completing
assessments for medical image analysis software \cite{Dong2021} and for Lattice
Boltzmann Method (LBM) software \cite{Michalski2021}.  The estimate assumes 30
software packages will be measured; the numbers will need to be adjusted if the
total packages changes.

@@

\begin{enumerate}
	\item The software functionality must fall within the identified domain.
	\item The source code must be viewable.
	\item The repository based measures should be available, which implies a
	preference for GitHub-style repositories.
	\item The software cannot be marked as incomplete or in an initial state of
	development.
\end{enumerate}

@@

If the list of software is too long (over around 30 packages), then steps need
to be taken to create a more manageable list to answer~\rqref{RQ_WhatProjects}.
The following filters were applied in the priority order listed. Copies of both
the initial and filtered lists, along with the rationale for shortening the
list, should be kept for traceability purposes.

@@

To answer \rqref{RQ_PainPoints}, we ask developers about their pain points and
compare their responses to pain points identified in the
literature~\cite{WieseEtAl2019,PintoEtAl2018}.  Pain points to watch for
include: lack of development time, Cross-platform compatibility, interruptions
while coding, scope bloat, lack of user feedback, hard to collaborate on
software projects, aloneness, dependency management, data handling concerns
(like data quality, data management and data privacy), reproducibility, and
software scope determination. For the LBM example, two sample pain points are:

\begin{enumerate}

	\item[P\refstepcounter{pnum}\thepnum \label{P_LackDevTime}:] \textbf{Lack of
	Development Time} A developer of pyLBM noted that their small development
	team has a lack of time to implement new features. Small development teams
	are common for LBM software packages (as shown in the measurement table
	excerpt in Figure~\ref{measurement_template_image}). Lack of time is also
	highlighted as a pain point by other research software developers
	\cite{PintoEtAl2018,WieseEtAl2019}.

	\item[P\refstepcounter{pnum}\thepnum \label{P_LackSoftDevExp}:] \textbf{Lack
	of Software Development Experience} A lack of software development
	experience was noted by the developer of TCLB, and others noted a need for
	improving software engineering education. Many of the team members on their
	project are domain experts, not computer scientists or software engineers.
	This same trend is noted by \cite{Nguyen-HoanEtAl2010}, which showed only
	23\% of research software survey respondents having a computing-related
	background. Similarly, \cite{UditAndKatz2017} show that the majority (54\%)
	of postdocs have not received training in software development.  The LBM
	developer suggesting an increasing role for formal software education
	matches the trend observed by \cite{PintoEtAl2018}, where their replication
	of a previous study \cite{HannayEtAl2009}, shows a growing interest in
	formal training (From 13\% of respondents in 2009 to 22\% in 2018).
	\cite{PintoEtAl2018} found that some developers feel there is a mismatch
	between coding skills and subject-matter skills.

\end{enumerate}

@@

Next step to recommend ways to improve software (previously section 9).  We wish
to know what practices are used by the top domain projects, so that others can
potentially emulate these practices. We also wish to identify new practices by
borrowing successful ideas from other domains.

Our comparison may point out areas where some LBM software
packages fall short of current best practices. This is not intended to be a
criticism of any existing packages, especially since in practice not every
project needs to achieve the highest possible quality.  However, rather than
delve into the nuances of which software can justify compromising which
practices we will write our comparison under the ideal assumption that every
project has sufficient resources to match best practices.

For each domain we wish to highlight success stories that can be shared amongst
domain community, and with the broader research software community, while at the
same time watching for areas for potential future improvement.

We have outlined a methodology for assessing the state of the practice for any
given research software domain.  When applying the methodology to a given domain, we provide a
means to answer the following questions:
\begin{inparaenum}[i)]
\item What artifacts (documents, code, test cases, etc.) are present?
\item What tools are used?
\item What principles, process and methodologies are used?
\item What are the pain points for developers?
\item What actions are used to improve qualities like maintainability and
reproducibility?
\item What specific actions are taken to achieve the qualities of usability,
traceability, modifiability, maintainability, correctness, understandability,
unambiguity, reproducibility and visibility/transparency?
\item How does software designated as high quality by this methodology compare
  with top rated software by the community?
\end{inparaenum}

The methodology depends on the engagement of a Domain Expert.  The Domain
Expert's role is to ensure that the assessment is consistent with the culture of
the community of practitioners in the domain.  The Domain Expert also has an
important role to play with the domain analysis.  For each domain we 
conduct a domain analysis to look at the commonalities, variabilities and
parameters of variation, for the family of software in the domain.  The domain
analysis means that software can be compared not just based on its quality, but
also based on its functionality.

The methodology follows a systematic procedure that begins with identifying the
domain and ends with answering the research questions posed above.  In between
we collect an authoritative list of about 30 software packages.  For each
package in the list we fill in our measurement template.  The template consists
of repository related data (like number of open issues, number of lines of code,
etc.) and 108 measures/questions related to 9 qualities: installability,
correctness/verifiability, reliability, robustness, usability, maintainability,
reusability, understandability and visibility/transparency. Filling in the
template requires installing the software, running simple tests (like completing
the getting started instructions (if present)), and searching the code,
documentation and test files.

The data for each domain is used to rank the software package according to each
quality dimension using AHP.  The ranking is not intended to identify a single
best software package.  Instead the ranking is intended to provide insights on
the top set of software for each quality.  The top performers can be contrasted
with the lesser performers to gain insight into what practices in the domain are
working.  Deeper insight can be obtained by combining this data with the
interview data from asking each recruited developer 20 questions.

Combining the quantitative data from the measurement template with the interview
results, along with the domain experts knowledge, we can determine the current
state of the practice for domain X.  Using our methodology, spreadsheet
templates and AHP tool, we estimate (based on our experience with using the
process) the time to complete an assessment for a given domain at 173 person
hours.

With the wealth of data from assessing the state of practice for multiple
domains, the next step is a meta-analysis.  We would look at how the different
domains compare. What lessons from one domain could be applied in other domains?
What (if any) differences exist in the pain points between domains?  Are there
differences in the tools, processes, and documentation between domains?

The goal of this assessment isn't to rank the software, but to use the ranking
exercise as a means to understand the state of the practice of LBM software
development.

@@

% \bibitem{ref_url1}
% LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
% Oct 2017
% \end{thebibliography}
\end{document}

>>>>

	We have developed a methodology for assessing the state of the software
	development practices for a given research software domain. Our methodology
	prescribes the following steps: i) Identify the domain; ii) Identify a list
	of candidate software packages; iii) Filter the list to a length of about 30
	packages; iv) Collect repository related data on each software package, like
	number of stars, number of open issues, number of lines of code; v) Fill in
	the measurement template (the template consists of 108 questions to assess 9
	qualities (including the qualities of installability, usability and
	visibility)); vi) Interview developers (the interview consists of 20
	questions and takes about an hour); and, vii) Rank the software using the
	Analytic Hierarchy Process (AHP). The data is analyzed to: i) comparing the
	ranking by best practices against an estimate of the communities ranking of
	popularity; ii) comparing artifacts, tools and processes to current research
	software development guidelines; and, iii) exploring pain points. We
	estimate the time to complete an assessment for a given domain at 173 person
	hours. We illustrate our method via the example of Lattice Boltzmann
	Solvers, where we find that the top packages engaged in most of recommended
	best practices, but show room for improvement with respect to providing API
	documentation, a roadmap, a code of conduct, programming style guide,
	uninstall instructions and continuous integration.

\keywords{research software \and software quality \and empirical measures \and analytic hierarchy process \and software artifacts \and developer pain points \and Lattice Boltzmann Method}

\end{abstract}

@@

Close to our goal of 160 person hours, we estimate 173 hours to complete the
assessment of a given domain~\cite{SmithEtAl2021}.  This estimate assumes the
domain has been decided, the Domain Expert has been recruited, and the template
spreadsheets and AHP tool~\cite{SmithEtAl2021}, will be utilized. The hours are
a rough estimate, based on our experience completing assessments for medical
image analysis~\cite{Dong2021} and the Lattice Boltzmann Method
(LBM)~\cite{Michalski2021}.  The estimate should be adjusted if the number of
packages measured differs from 30.

@@

\begin{enumerate}
	\item The software functionality must fall within the identified domain.
	\item The source code must be viewable.
	\item The repository based measures should be available, which implies a
	preference for GitHub-style repositories.
	\item The software cannot be marked as incomplete.
\end{enumerate}

@@

If the list of software is too long (over around 30 packages)
for~\rqref{RQ_WhatProjects}, then filters are applied in the priority order
listed. Copies of both the initial and filtered lists, along with the rationale
for shortening the list, should be kept for traceability purposes.

@@

To answer \rqref{RQ_PainPoints}, we ask developers about their pain points and
compare their responses to pain points identified in the
literature~\cite{WieseEtAl2019,PintoEtAl2018}.  Pain points to watch for
include: lack of development time, Cross-platform compatibility, interruptions
while coding, scope bloat, lack of user feedback, hard to collaborate on
software projects, aloneness, dependency management, data handling concerns
(like data quality, data management and data privacy), reproducibility, and
software scope determination.

An example pain point noted by LBM developers is a lack of development time. A
developer of pyLBM noted that their small development team did not have enough
time to implement new features. Small development teams are common for LBM
software packages (as shown in the measurement table excerpt in
Figure~\ref{measurement_template_image}).

@@

To improve the development of research software, both in terms of productivity,
and the resulting software quality, we need to understand the current state of
the practice. An exciting strategy to approach this goal is to assess one domain
at a time, collecting data from developer interviews and digging deeply into
their code repositories. By providing feedback specific to their domain, the
developer community can be drawn into a dialogue on how to best make
improvements going forward. Moreover, they can be encouraged to share their best
practices between one another, and with other research software domains.

We have outlined a methodology for assessing the state of the practice for any
given research software domain based on a list of about 30 representative
projects.  In addition to interviews, we use manual and automated inspection of
the artifacts in each project's repositories.  The repository data is collected
by filling in a 108 question measurement template, which requires installing the
software on a VM, running simple tests (like completing the getting started
instructions (if present)), and searching the code, documentation and test
files. Using AHP the projects are ranked for 9 individual qualities
(installability, correctness and verifiability, reliability, robustness,
usability, maintainability, reusability, surface understandability,
visibility/transparency) and for overall quality.  Perspective and insight is
shared with the user community via the following: i) comparing the ranking by
best practices against an estimate of the communities ranking of popularity; ii)
comparing artifacts, tools and processes to current research software
development guidelines; and, iii) exploring pain points via developer
interviews. Using our methodology, spreadsheet templates and AHP tool, we
estimate (based on our experience with using the process) the time to complete
an assessment for a given domain at 173 person hours.

For the running example of LBM we found that the top packages engaged in most of
recommended best practices, including examples of practising peer review.
However, we did find room for improvement with respect to providing API
documentation, a roadmap, a code of conduct, programming style guide, and
uninstall instructions.  In addition, the community could likely benefit by
increased use of continuous integration.

Building from the LBM example we can create a wealth of data on multiple
domains.  The next step is a meta-analysis, where we look at how the different
domains compare to answer new research questions like: What lessons from one
domain could be applied in other domains? What (if any) differences exist in the
pain points between domains?  Are there differences in the tools, processes, and
documentation between domains?

@@

% \bibitem{ref_url1}
% LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
% Oct 2017
% \end{thebibliography}

\end{document}
